Remote sensing the gas exchange velocity and carbon efflux using the SWOT Satellite

### 2021-02-03 manuscript_v1.Rmd
ms updates

---
title: "Remote sensing the gas exchange velocity and carbon efflux using the SWOT Satellite"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##1 Introduction
Greenhouse gas (GHG) emissions from rivers and lakes are significant (Cole et al. 2007). Rivers are usually supersaturated with dissolved GHGs and present a net flux of these gases from water to air (Cole & Caraco 2001; Jones et al. 2003). This flux $[M/L^2T]$ is calculated, with in situ knowledge of the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$, as equation 1.

$Flux=k(C_{water}-C_{air})$ (1)

k is largely a function of water column turbulence, at least in fluvial settings. This is because k scales with diffusivity as normalized by the boundary-layer thickness, and in rivers turbulence defines the thickness of the boundary-layer (Hall & Ulseth, 2020 for a thorough review of gas exchange velocity in rivers). Due to its reliance on diffusivity (as well as kinematic viscosity), k is defined differently for different GHGs (Raymond et al. 2012; Wanninkof 1992). It is also often measured using different gases than those being studied, so a variant of k that is normalized to a Schmidt number of 600 ($k_{600} [L/T]$) is generally used to be GHG and temperature agnostic (Hall & Ulseth, 2020). This is synonymous with a k for carbon dioxide ($k_{CO2}$) at 20 degrees Celsius.

It is infeasible to measure $k_{600}$ directly when working at the river network scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology (e.g. O’Connor & Dobbins, 1958; Palumbo & Brown, 2014; Raymond et al. 2012). Following the theory above, many upscaling functions are based on the basic concept that k is empirically correlated with the turbulent energy dissipation e at the air-water interface (Zappa et al. 2007). This is convenient for upscaling because in rivers and streams e can be equated to the turbulent energy dissipation rate eD $[L^2/T^3]$ (Ulseth et al. 2019; Raymond et al. 2012), which is estimated using easily calculated hydraulic parameters (eD≈gSV where g is gravitational acceleration, S is channel slope, and V is average flow velocity- Tsivoglou & Neal, 1976). It has been observed that this scaling relationship is particularly strong in higher energy streams where bubble-mediated gas exchange occurs, though the general scaling function holds across all rivers and streams (Ulseth et al. 2019). Despite reasonable predictive accuracy, these upscaling models suffer from residual variation indicating large predictive error (Wallin et al. 2011). Still, these are the current best options for estimating $k_{600}$ in rivers. It has been suggested that exploring at-a-station hydraulic variation (and thus $k_{600}$ variation) is a promising approach to reducing the residual variation in upscaling models (Hall & Ulseth, 2020).

Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to river networks are limited by estimates of the reach-averaged flow velocity of the river channel. Therefore, network-scale GHG flux estimates exist in some part at the mercy of the quality of average flow velocity data. Ideally, a worker can directly measure velocity at some permanently or temporarily gauged river in their system and build velocity~discharge rating curves to upscale to the entire river network. Rating curves are also referred to as hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge (Leopold & Maddock, 1953). However, this is not feasible in ungauged basins (whether physically or politically ungauged- Gleason & Durand, 2020) where no in situ hydraulic information is available and fieldwork is impractical. Thus, in ungauged settings, present understandings of GHG fluxes from rivers are likely limited by our access to velocity and/or discharge data.

To circumvent this problem in ungauged basins/at massive spatial scales, researchers have used ‘global-scope’ HG models that use a single set of model parameters trained on large hydraulics datasets. This is the approach used in upscaling average flow velocity to continental/global river networks and has proven quite useful in those settings (e.g. Horgby et al. 2019; Borges et al. 2015; Lauerwald et al. 2015; Raymond et al. 2013). However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river (Park, 1977). Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters (e.g. Dingman 2007; Parker, et al. 2007; Singh, 2003; Ferguson, 1986) and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). A potential alternative to this approach, in ungauged basins at least, is to directly estimate a river’s hydraulic properties from remote sensing (RS) data.

The problem of hydraulic estimation in ungauged basins is itself a burgeoning subfield within fluvial hydrology and an analogue to the problem presented here is the ungauged remote sensing of river discharge (RSQ- see Gleason & Durand, 2020 for a thorough review). A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g. Andreadis et al. 2020; Brinkerhoff et al. 2020a; Durand et al. 2014; Garambois & Monnier, 2015; Hagemann et al. 2017). They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able (Gleason & Durand, 2020; Gleason et al. 2017). McFLIs have largely been developed in the context of the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) mission, which will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions (Biancamaria et al. 2016). On average SWOT will sample rivers every 11 days and provide a unique opportunity to measure river and lake hydraulics at the global scale. While the raw performance of McFLIs is lower than a more sophisticated approach, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our understandings of ungauged rivers when little to no information was previously available (Brinkerhoff et al. 2020a; Durand et al. 2016). Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation (Ishitsuka et al. 2020).

McFLIs often employ Bayesian inference. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen (O2) datasets (Appling et al. 2018; Grace et al. 2015; Holtgrieve et al. 2010). While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian inference could be useful for estimating $k_{600}$ when parameter equifinality is a problem (at it is with McFLIs).

@@

###2.1 Developing an RS-able $k_{600}$ upscaling model
Raymond et al. (2012) first used river hydraulics and the theoretical relationship between the turbulent dissipation rate (eD) and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g. O’Connor & Dobbins, 1958; Tsivoglou & Neal, 1976). Later, Ulseth et al. (2019) built upon Raymond et al. (2012)’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, Ulseth et al. (2019) scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether eD is high or low. Their upscaling function is reprinted as equation 2.

$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02$  (2)

$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of eD to assign its differential model parameters. We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the Ulseth et al. (2019) model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the Ulseth et al. 2019 dataset). In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

$k_{600}=a_i(eD)^{b_i}$ (3)

This model therefore assigns ai and bi a priori using just river width and slope to define i. Another implication of this extension of the Ulseth et al. (2019) model is that we now use a and b parameters that implicitly reflect the distinct geomorphology of different sized rivers.
Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original Ulseth et al. (2019) model, with the goal of at least replicating their performance. We compared it against the various eD-based upscaling models from Raymond et al. (2012) that are commonly used in the literature. This was done by training all upscaling functions on 80% of the Ulseth et al. (2019) data and comparing their performance on the remaining 20% of the data. Because the Raymond et al. (2012) models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of eD (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$. The approach used here is heavily informed by Hagemann et al. (2017)’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize upscaling $k_{600}$ as a Bayesian inference problem.

@@

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following Tsivoglou & Neal (1976), $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width W [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area dA). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i}$  (4)

@@

$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})$  (5)

Equation five necessitates that Bayesian priors be assigned to model parameters n, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively, they represent our ‘prior river knowledge’ of what n, A0, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean (μ), standard deviation (σ), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. n and $A_0$ prior hyperparameters were assigned following the method developed by Brinkerhoff et al. (2020a). They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the Ulseth et al. (2019) dataset to assign μ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634$  (6)

$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Uncertainty in equation 4 can stem from two sources: 1) error baked into the upscaling model (ai and bi parameter uncertainty) and 2) error in Manning’s approximation of the average flow velocity. To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by Brinkerhoff et al. (2019). Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, and ln-transformed velocity. Estimate uncertainty for velocity followed Hagemann et al. (2017), who estimated ln-transformed Manning’s equation error to be 0.25. This resulted in 8,000 discrete distributions for $k_{600}$ from which the average uncertainty term is extracted and used for $σ_{upscale}^2$.

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces errors to these simulated rivers (mostly from radar layover error and random noise- Oubanas et al. 2018a). Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no observation errors) and represent the best-case scenarios of what SWOT will provide to hydrologists. Here, we use the 22 rivers archived in Frasson et al. (2019) and used by Rodriquez et al. (2020) to develop reach-averaged Saint Vernant equations within the SWOT context. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly CO2 data used later- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

@@

$k_{600, obs}=a_i(gS_hV_{obs})^{b_i}$ (7)

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination r2 and the root mean square error (RMSE). Four normalized metrics were used for by-river validation to compare across rivers: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value (Knoben et al. 2019) and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.

@@

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations. 29 bi-weekly $CO_2$ samples were made by Beaulieu et al. (2012- Figure S1) at one location in the Ohio River. We paired these 29 values with each 11-day-repeat set of SWOT observations, ignoring the timesteps beyond 29 (only 7% of the SWOT observations were ignored here and this was deemed acceptable). This $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm and water temperature was assumed 25 degrees Celsius. The Schmidt number, used to obtain kCO2 from $k_{600}$, was calculated following Raymond et al. (2012) and Wanninkof et al. (1992).

Then, we calculated bulk carbon efflux using four models for average flow velocity:  BIKER and three rating curves previously used for $CO_2$ upscaling (Lauerwald et al. 2015; Raymond et al. 2013; Raymond et al. 2012- see Table S1 for their definitions). It is worth stressing that the Lauerwald, et al. (2015) model was developed as one of two components of the Raymond et al. (2013) velocity model, and so was actually developed by the latter’s authors. The names used here refer to the final implementation used in each study’s upscaling exercises. For this study, all rating-curve methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method versus two gauged methods (that also use global-scope HG parameters). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) [M/T] as the average mass flow rate of carbon per year, i.e. multiplied by the surface area of all 22 rivers.

>>>>

---
title: "Remote sensing the gas exchange velocity and carbon efflux using the SWOT Satellite"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
bibliography: References.bib
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##1 Introduction
Greenhouse gas (GHG) emissions from rivers and lakes are significant [@colePlumbingGlobalCarbon2007a]. Rivers are usually supersaturated with dissolved GHGs and present a net flux of these gases from water to air [@coleCarbonCatchmentsConnecting2001a; @jonesLongtermDeclineCarbon2003]. This flux $[M/L^2T]$ is calculated, with in situ knowledge of the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$, as equation 1.

$$Flux=k(C_{water}-C_{air}) (1)$$

k is largely a function of water column turbulence, at least in fluvial settings. This is because k scales with diffusivity as normalized by the boundary-layer thickness, and in rivers turbulence defines the thickness of the boundary-layer (@hallGasExchangeStreams2020 for a thorough review of gas exchange velocity in rivers). Due to its reliance on diffusivity (as well as kinematic viscosity), k is defined differently for different GHGs [@raymondScalingGasTransfer2012a; @wanninkhofRelationshipWindSpeed1992a]. It is also often measured using different gases than those being studied, so a variant of k that is normalized to a Schmidt number of 600 ($k_{600} [L/T]$) is generally used to be GHG and temperature agnostic [@hallGasExchangeStreams2020]. This is synonymous with a k for carbon dioxide ($k_{CO2}$) at 20 degrees Celsius.

It is infeasible to measure $k_{600}$ directly when working at the river network scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Following the theory above, many upscaling functions are based on the basic concept that k is empirically correlated with the turbulent energy dissipation e at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams e can be equated to the turbulent energy dissipation rate eD $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters (eD≈gSV where g is gravitational acceleration, S is channel slope, and V is average flow velocity- @tsivoglouTracerMeasurementReaeration1976). It has been observed that this scaling relationship is particularly strong in higher energy streams where bubble-mediated gas exchange occurs, though the general scaling function holds across all rivers and streams [@ulsethDistinctAirWater2019a]. Despite reasonable predictive accuracy, these upscaling models suffer from residual variation indicating large predictive error (Wallin et al. 2011). Still, these are the current best options for estimating $k_{600}$ in rivers. It has been suggested that exploring at-a-station hydraulic variation (and thus $k_{600}$ variation) is a promising approach to reducing the residual variation in upscaling models [@hallGasExchangeStreams2020].

Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to river networks are limited by estimates of the reach-averaged flow velocity of the river channel. Therefore, network-scale GHG flux estimates exist in some part at the mercy of the quality of average flow velocity data. Ideally, a worker can directly measure velocity at some permanently or temporarily gauged river in their system and build velocity~discharge rating curves to upscale to the entire river network. Rating curves are also referred to as hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a]. However, this is not feasible in ungauged basins (whether physically or politically ungauged- @gleasonRemoteSensingRiver2020) where no in situ hydraulic information is available and fieldwork is impractical. Thus, in ungauged settings, present understandings of GHG fluxes from rivers are likely limited by our access to velocity and/or discharge data.

To circumvent this problem in ungauged basins/at massive spatial scales, researchers have used ‘global-scope’ HG models that use a single set of model parameters trained on large hydraulics datasets. This is the approach used in upscaling average flow velocity to continental/global river networks and has proven quite useful in those settings (e.g. @horgbyUnexpectedLargeEvasion2019; @borgesVariabilityGasTransfer2004; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a). However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977]. Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters (e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a) and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). A potential alternative to this approach, in ungauged basins at least, is to directly estimate a river’s hydraulic properties from remote sensing (RS) data.

The problem of hydraulic estimation in ungauged basins is itself a burgeoning subfield within fluvial hydrology and an analogue to the problem presented here is the ungauged remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; HIVDI; @hagemannBAMBayesianAMHGManning2017a). They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. McFLIs have largely been developed in the context of the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) mission, which will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions [@biancamariaSWOTMissionIts2016]. On average SWOT will sample rivers every 11 days and provide a unique opportunity to measure river and lake hydraulics at the global scale. While the raw performance of McFLIs is lower than a more sophisticated approach, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote].

McFLIs often employ Bayesian inference. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen (O2) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian inference could be useful for estimating $k_{600}$ when parameter equifinality is a problem (at it is with McFLIs).

@@

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between the turbulent dissipation rate (eD) and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976). Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether eD is high or low. Their upscaling function is reprinted as equation 2.

$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  (2) \\

ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of eD to assign its differential model parameters. We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the [@ulsethDistinctAirWater2019a] dataset). In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

$$k_{600}=a_i(eD)^{b_i} (3)$$

This model therefore assigns ai and bi a priori using just river width and slope to define i. Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use a and b parameters that implicitly reflect the distinct geomorphology of different sized rivers.

Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original [@ulsethDistinctAirWater2019a] model, with the goal of at least replicating their performance. We compared it against the various eD-based upscaling models from [@raymondScalingGasTransfer2012a] that are commonly used in the literature. This was done by training all upscaling functions on 80% of the [@ulsethDistinctAirWater2019a] data and comparing their performance on the remaining 20% of the data. Because the [@raymondScalingGasTransfer2012a] models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of eD (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$. The approach used here is heavily informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize upscaling $k_{600}$ as a Bayesian inference problem.

@@

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following [@tsivoglouTracerMeasurementReaeration1976], $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width W [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area dA). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} (4)$$

@@

$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  (5)$$

Equation five necessitates that Bayesian priors be assigned to model parameters n, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively, they represent our ‘prior river knowledge’ of what n, A0, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean (μ), standard deviation (σ), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. n and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020]. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign μ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   (6) \\

\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Uncertainty in equation 4 can stem from two sources: 1) error baked into the upscaling model (ai and bi parameter uncertainty) and 2) error in Manning’s approximation of the average flow velocity. To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by [@brinkerhoffReconcilingAtaStationAtManyStations2019]. Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, and ln-transformed velocity. Estimate uncertainty for velocity followed [@hagemannBAMBayesianAMHGManning2017a], who estimated ln-transformed Manning’s equation error to be 0.25. This resulted in 8,000 discrete distributions for $k_{600}$ from which the average uncertainty term is extracted and used for $σ_{upscale}^2$.

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces errors to these simulated rivers (mostly from radar layover error and random noise- @oubanasDischargeEstimationUngauged2018). Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no observation errors) and represent the best-case scenarios of what SWOT will provide to hydrologists. Here, we use the 22 rivers archived in [@frassonCompilationHydraulicModels2019a] and used by [@rodriguezObservingRiversVarying2020] to develop reach-averaged Saint Vernant equations within the SWOT context. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly CO2 data used later- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

@@

$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} (7)$$

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination r2 and the root mean square error (RMSE). Four normalized metrics were used for by-river validation to compare across rivers: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a] and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.

@@

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations. 29 bi-weekly $CO_2$ samples were made by [@beaulieuControlsGasTransfer2012a]- Figure S1) at one location in the Ohio River. We paired these 29 values with each 11-day-repeat set of SWOT observations, ignoring the timesteps beyond 29 (only 7% of the SWOT observations were ignored here and this was deemed acceptable). This $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm and water temperature was assumed 25 degrees Celsius. The Schmidt number, used to obtain kCO2 from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].

Then, we calculated bulk carbon efflux using four models for average flow velocity:  BIKER and three rating curves previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondScalingGasTransfer2012a] velocity model, and so was actually developed by the latter’s authors. The names used here refer to the final implementation used in each study’s upscaling exercises. For this study, all rating-curve methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method versus two gauged methods (that also use global-scope HG parameters). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) [M/T] as the average mass flow rate of carbon per year, i.e. multiplied by the surface area of all 22 rivers.
